---
title: "deseq_571"
output: html_document
date: "2024-12-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# preprocessing to obtain counts and conditions
```{r}
library(tidyverse)

file_path <- "osdr571.csv"
data <- read.csv(file_path, sep = ";", check.names = FALSE)

# data <- data %>% select(-`Unnamed: 26`)
rownames(data) <- data$`Unnamed: 0`
# data <- data %>% select(-`Unnamed: 0`)

l_samples <- colnames(data)[grepl("_L", colnames(data))]
r_samples <- colnames(data)[grepl("_R", colnames(data))]

counts_matrix <- data[, c(l_samples, r_samples)]

sample_conditions <- data.frame(
  sample = c(l_samples, r_samples),
  condition = c(rep("L", length(l_samples)), rep("R", length(r_samples)))
)
write.csv(counts_matrix, "counts_matrix.csv", row.names = TRUE)
write.csv(sample_conditions, "sample_conditions.csv", row.names = FALSE)
```

# load packages and needed data
```{r}
library(DESeq2)

count_data <- read.csv("counts_matrix.csv", row.names = 1)
sample_data <- read.csv("sample_conditions.csv", row.names = 1)
```

# create dds object
```{r message=FALSE, warning=FALSE}
dds <- DESeqDataSetFromMatrix(
    countData = count_data,
    colData = sample_data,
    design = ~ condition
)
```

# run dds and extract the results
```{r}
dds <- DESeq(dds)
```


```{r}
results <- results(dds, contrast = c("condition", "R", "L"))

head(results)
write.csv(as.data.frame(results), file = "deseq571_results.csv")

```

# plot the results
```{r}
plotMA(results, ylim = c(-2, 2))
DESeq2::plotPCA(vst(dds))

library(ggplot2)
results <- read.csv("deseq571_results.csv", row.names = 1)
log2fc_threshold <- 1
pval_threshold <- 0.05

results$significant <- ifelse(
  results$padj < pval_threshold & abs(results$log2FoldChange) > log2fc_threshold,
  "Significant", "Not Significant"
)

results$direction <- ifelse(
  results$significant == "Significant" & results$log2FoldChange > 0, "Upregulated",
  ifelse(results$significant == "Significant" & results$log2FoldChange < 0, "Downregulated", "Not Significant")
)

ggplot(results, aes(x = log2FoldChange, y = -log10(pvalue), color = direction)) +
  geom_point(alpha = 0.8, size = 2) +
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Not Significant" = "gray")) +
  theme_minimal() +
  labs(
    title = "Volcano Plot",
    x = "Log2 Fold Change",
    y = "-Log10 P-value",
    color = "Gene"
  ) +
  theme(legend.position = "top")
```

# assign the genes name to the results
```{r}
osdr571 <- read.csv("osdr571.csv", header = TRUE, row.names = 1)
deseq_results <- read.csv("deseq571_results.csv", header = TRUE)

if (nrow(deseq_results) == nrow(osdr571)) {
     rownames(deseq_results) <- rownames(osdr571)
     write.csv(deseq_results, "deseq571_results_with_genes.csv", row.names = TRUE)
     head(deseq_results)
 } else {
    cat("Warning: The number of rows in the two files do not match.\n")
   }
```

# bit of post-processing to obtain final file
```{r}
deseq_results <- read.csv("deseq571_results_with_genes.csv", header = TRUE, row.names = 1)

deseq_results <- deseq_results %>% dplyr::select(-X)

cleaned_rownames <- sub("\\..*", "", rownames(deseq_results))
cleaned_rownames <- make.unique(cleaned_rownames)
rownames(deseq_results) <- cleaned_rownames

write.csv(deseq_results, "deseq571_results_averageVSaverage.csv", row.names = TRUE)

head(deseq_results)
```
# check for lncrnas
```{r}
linclist <- read.delim("linclist.tsv", header = T)
lncRNAs <- deseq_results %>%
  dplyr::filter(rownames(deseq_results) %in% linclist$ensemblGeneID)
```

# make a volcano to visualize expression of DE lncrnas (average vs average)
```{r}
ggplot(lncRNAs, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(aes(color = ifelse(padj < 0.05 & abs(log2FoldChange) > 1, 
                                ifelse(log2FoldChange > 0, 'Upregulated', 'Downregulated'), 'Not Significant')),
             alpha = 0.8, size = 2) +
  scale_color_manual(values = c('Upregulated' = 'red', 'Downregulated' = 'blue', 'Not Significant' = 'gray')) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  labs(title = "Volcano Plot for lncRNAs", 
       x = "log2(Fold Change)", 
       y = "-log10(P-value)",
       color = "lncRNAs") +
  theme_minimal() +
  theme(legend.position = "top")
```
# extract the most significant lncrnas, up and down
```{r}
upregulated_lncRNAs <- lncRNAs %>%
  filter(log2FoldChange > 1 & pvalue < 0.1) %>%
  rownames()

downregulated_lncRNAs <- lncRNAs %>%
  filter(log2FoldChange < -1 & pvalue < 0.1) %>%
  rownames()

upregulated_pvals <- lncRNAs %>%
  filter(rownames(lncRNAs) %in% upregulated_lncRNAs) %>%
  pull(pvalue) %>%
  sort()
plot(upregulated_pvals, main = "Sorted p-values for Upregulated lncRNAs", 
     xlab = "Gene Index", ylab = "p-value", type = "b", col = "red")

downregulated_pvals <- lncRNAs %>%
  filter(rownames(lncRNAs) %in% downregulated_lncRNAs) %>%
  pull(pvalue) %>%
  sort()
plot(downregulated_pvals, main = "Sorted p-values for Downregulated lncRNAs", 
     xlab = "Gene Index", ylab = "p-value", type = "b", col = "blue")
```
# filter the whole matrix, not only rownames... and save the results
```{r}
upregulated_lncRNAs_df <- lncRNAs %>%
  filter(log2FoldChange > 1 & pvalue < 0.1) 

downregulated_lncRNAs_df <- lncRNAs %>%
  filter(log2FoldChange < -1 & pvalue < 0.1)

filtered_lncRNAs_df <- bind_rows(upregulated_lncRNAs_df, downregulated_lncRNAs_df)
write.table(filtered_lncRNAs_df, "lncRNAs_571_averVSaver.tsv", sep = "\t", quote = FALSE, row.names = TRUE)

head(filtered_lncRNAs_df)
```
# add a gene symbol columnm
```{r}
library(biomaRt)

ensembl <- useMart("ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl")

gene_mapping <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                      filters = "ensembl_gene_id",
                      values = rownames(filtered_lncRNAs_df), 
                      mart = ensembl)

filtered_lncRNAs_with_symbols <- merge(filtered_lncRNAs_df, gene_mapping, 
                                       by.x = "row.names", by.y = "ensembl_gene_id", 
                                       all.x = TRUE)

filtered_lncRNAs_with_symbols$gene_symbol <- filtered_lncRNAs_with_symbols$hgnc_symbol

head(filtered_lncRNAs_with_symbols)

write.table(filtered_lncRNAs_with_symbols, "lncRNAs_571_averVSaver_symbols.tsv", sep = "\t", quote = FALSE, row.names = TRUE)
```
# retrieve the targets of the lncRNAs
```{r}
top10list <- read.csv("top10list_average.csv", row.names = 1, stringsAsFactors = FALSE)
lnc2target <- read.csv("lnc2target.csv", stringsAsFactors = FALSE)

matched_targets <- lnc2target %>%
  filter(Ensembl_ID %in% rownames(top10list))

result <- matched_targets %>%
  dplyr::select(Ensembl_ID, Target_entrez_gene_ID)

# write.table(result, "matched_targets.tsv", sep = "\t", quote = FALSE, row.names = FALSE)

head(result)

```
# plotting the results obtained on disgenet inserting the most 10 de lncrnas,
# 5 up and 5 down. qui fatto solo per average, da fare per gli altri.
```{r}
data <- read.delim("average_disgenet.tsv", sep = "\t")


data <- data %>%
  mutate(GeneList = strsplit(as.character(Gene), ",")) %>%
  unnest(GeneList) %>%
  distinct(Disease, GeneList) %>%
  group_by(Disease) %>%
  summarise(GeneCount = n())  

disease_gene_counts <- data %>%
  arrange(-GeneCount)

ggplot(disease_gene_counts, aes(x = GeneCount, y = reorder(Disease, GeneCount))) +
  geom_bar(stat = "identity", fill = "darkred") +
  labs(
    #title = "lncRNAs-Diseases associations",
    x = "Number of lncRNAs",
    y = "Disease"
  ) +
  scale_x_continuous(
    limits = c(0, 3),             
    breaks = seq(0, 3, by = 1)     
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    #plot.title = element_text(hjust = 0.01)
  ) 
```

# doing the same for long term
# qua da notare che sono in grado di recuperare alcuni targets dei most DE lncrna longterm!!!
# cosa che non era stata possibile per average :/
```{r}
top10list <- read.csv("top10list_long_term.csv", row.names = 1, stringsAsFactors = FALSE)
lnc2target <- read.csv("lnc2target.csv", stringsAsFactors = FALSE)

matched_targets <- lnc2target %>%
  filter(Ensembl_ID %in% rownames(top10list))

result <- matched_targets %>%
  dplyr::select(Ensembl_ID, Target_entrez_gene_ID)

# write.table(result, "matched_targets.tsv", sep = "\t", quote = FALSE, row.names = FALSE)

head(matched_targets)
```

# plotting the results obtained on disgenet inserting the most 10 de lncrnas,
# 5 up and 5 down. qui fatto solo per long term.
```{r}
data <- read.delim("long_term_disgenet.tsv", sep = "\t")

data <- data %>%
  mutate(GeneList = strsplit(as.character(Gene), ",")) %>%
  unnest(GeneList) %>%
  distinct(Disease, GeneList) %>%
  group_by(Disease) %>%
  summarise(GeneCount = n())  

disease_gene_counts <- data %>%
  arrange(-GeneCount)

ggplot(disease_gene_counts, aes(x = GeneCount, y = reorder(Disease, GeneCount))) +
  geom_bar(stat = "identity", fill = "darkred") +
  labs(
    #title = "lncRNAs-Diseases associations",
    x = "Number of lncRNAs",
    y = "Disease"
  ) +
  scale_x_continuous(
    limits = c(0, 2),             
    breaks = seq(0, 2, by = 1)     
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    #plot.title = element_text(hjust = 0.01)
  )
```

# doing the same for short term
# qua da notare che NON sono in grado di recuperare alcuni targets dei most DE lncrna longterm 
# cosa uguale ad average :/
```{r}
top10list <- read.csv("top10list_short_term.csv", row.names = 1, stringsAsFactors = FALSE)
lnc2target <- read.csv("lnc2target.csv", stringsAsFactors = FALSE)

matched_targets <- lnc2target %>%
  filter(Ensembl_ID %in% rownames(top10list))

result <- matched_targets %>%
  dplyr::select(Ensembl_ID, Target_entrez_gene_ID)

# write.table(result, "matched_targets.tsv", sep = "\t", quote = FALSE, row.names = FALSE)

head(matched_targets)
```

# plotting the results obtained on disgenet inserting the most 10 de lncrnas,
# 5 up and 5 down. qui fatto solo per short term.
```{r}
data <- read.delim("short_term_disgenet.tsv", sep = "\t")

data <- data %>%
  mutate(GeneList = strsplit(as.character(Gene), ",")) %>%
  unnest(GeneList) %>%
  distinct(Disease, GeneList) %>%
  group_by(Disease) %>%
  summarise(GeneCount = n())  

disease_gene_counts <- data %>%
  arrange(-GeneCount)

ggplot(disease_gene_counts, aes(x = GeneCount, y = reorder(Disease, GeneCount))) +
  geom_bar(stat = "identity", fill = "darkred") +
  labs(
    #title = "lncRNAs-Diseases associations",
    x = "Number of lncRNAs",
    y = "Disease"
  ) +
  scale_x_continuous(
    limits = c(0, 3),             
    breaks = seq(0, 3, by = 1)     
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    #plot.title = element_text(hjust = 0.01)
  )
```
# WCGNA analysis
```{r}
library(dplyr)

pre_samples <- readLines("lista_pre.txt")
post_samples <- readLines("lista_post.txt")
pre_df <- data.frame(Sample = pre_samples, Group = "Pre", stringsAsFactors = FALSE)
post_df <- data.frame(Sample = post_samples, Group = "Post", stringsAsFactors = FALSE)

combined_df <- bind_rows(pre_df, post_df)

write.csv(combined_df, "samples.csv", row.names = FALSE)

df <- read.csv("wgcna.csv")
linclist_genes <- linclist$ensemblGeneID
filtered_df <- subset(df, gene %in% linclist_genes)
write.csv(filtered_df, "wgcna_lncRNAs.csv", row.names = FALSE)
```
